Using Visual Analytics to Explain\\[1ex] Black-Box Machine Learning

### 2018-04-06 tex/introduction/motivation.tex
typo fixing

@@

\par \noindent (1)
Caruana~\etal\cite{Caruana:2015:IMH:2783258.2788613} built a Generalized Additive Model with pairwise interaction ($GA^2M$) to predict mortality risk for hospitalized patients with pneumonia, a potentially life-threatening lung infection. This area of machine learning is called predictive modeling and uses machine learning to predict a certain outcome, in this case mortality risk, based on values of several input features, in this case laboratory measurements, comorbidities, or patient demographics, and is trained from and applied to many instances. It is not feasible to individually check each prediction of the model manually. However, the machine learning algorithm $GA^2M$ used for this study is an intelligible model, that sacrifices parts of its predictive quality in favor of being understandable by humans. This enabled the authors to inspect how features contributed to the predicted outcome. One of their findings was that the model associated having asthma, a chronic lung disease, with a lower mortality rate for pneumonia patients. However, this combination of diseases is known to have a significantly \emph{increased} mortality rate. The authors double checked their data and found that this phenomenon actually occurred in their dataset and thus appeared as correct prediction in their model. After further investigation it became clear that this bias in the dataset stemmed from patients, with those conditions, being treated with intensive care thus lowering their mortality risk below the average. Outside of a hospital the risk of those patients would have been significantly higher. The model, however, only learnt about hospitalized patients which led to it being confident in this false association.

@@

The above examples illustrate the need for human understanding and supervision in machine learning.
However, those cases required intricate understanding of the used machine learning models and are not easily generalized to other techniques.
For example, in the case of Caruana~\etal\cite{Caruana:2015:IMH:2783258.2788613}, the incorrect correct result was easily identified since the authors used a special, intelligible, model.
However, not all models used in machine learning are, or even attempt to be, easily interpretable by human experts.
To counteract this, visual analytics has been successfully used to visualize machine learning processes in order to understand and gain insights into machine learning models~\cite{DBLP:journals/corr/StrobeltGHPR16,DBLP:journals/corr/abs-1710-06501,2017arXiv170401942K,DBLP:journals/corr/abs-1710-10777,8019879,8019872,DBLP:journals/tvcg/LiuXLWWZ18,8019878}. Visual analytics is the area of studying interactive graphical representations of abstract information with the help of computational and statistical procedures, in order to effectively analyze complex data. With regards to understanding the decision making of predictive machine learning models, there are two main strategies: white-box and black-box analysis \cite{class_signatures}.

>>>>

\par \noindent (1)
Caruana~\etal\cite{Caruana:2015:IMH:2783258.2788613} built a Generalized Additive Model with pairwise interaction ($GA^2M$) to predict mortality risk for hospitalized patients with pneumonia, a potentially life-threatening lung infection. This predictive modeling task used machine learning to predict the mortality risk based on values of several input features, \eg, laboratory measurements, comorbidities, or patient demographics, and was trained from and tested against many instances. It was not feasible to individually check each prediction of the model manually. However, the machine learning algorithm $GA^2M$ used for this study is an intelligible model, that sacrifices parts of its predictive quality in favor of being understandable by humans. This enabled the authors to inspect how features contributed to the predicted outcome.

One of their findings was that the model associated having asthma, a chronic lung disease, with a lower mortality rate for pneumonia patients. However, this combination of diseases is known to have a significantly \emph{increased} mortality rate.
The authors double checked their data and found that this phenomenon actually occurred in their data set and thus appeared as correct predictions in their model. After further investigation, it became clear that this bias in the data set stemmed from patients, with those conditions, being treated with intensive care thus lowering their mortality risk below the average. Outside of a hospital the risk of those patients would have been significantly higher. The model, however, only learnt about hospitalized patients which led to it being confident in this false association.

@@

@@

The above examples illustrate the need for human understanding and supervision in machine learning.
However, those cases required intricate understanding of the used machine learning models and are not easily generalized to other techniques.
For example, in the case of Caruana~\etal\cite{Caruana:2015:IMH:2783258.2788613}, the incorrect correct result was easily identified since the authors used a special, intelligible, model.
However, not all models used in machine learning are, or even attempt to be, easily interpretable by human experts.
To counteract this, visual analytics has been successfully used to visualize machine learning processes in order to understand and gain insights into machine learning models~\cite{DBLP:journals/corr/StrobeltGHPR16,DBLP:journals/corr/abs-1710-06501,2017arXiv170401942K,DBLP:journals/corr/abs-1710-10777,8019879,8019872,DBLP:journals/tvcg/LiuXLWWZ18,8019878}. 
% Visual analytics is the area of studying interactive graphical representations of abstract information with the help of computational and statistical procedures, in order to effectively analyze complex data.
With regards to understanding the decision making of predictive machine learning models, there are two main strategies: white-box and black-box analysis \cite{class_signatures}.

### 2018-04-06 tex/summary.tex
typo fixing

In this thesis, we discussed the role of visual analytics to explain black-box machine learning.
We saw that global approaches, like feature selection methods, are not informative enough to help understand black-box model decisions.
Different strategies preferred different, equally reasonable, feature sets without having a significant impact on predictive performance.
This showed, that inspecting and comparing alternate settings let machine learning experts develop insights that overwrote their initial intuitions.

Next, we saw that partial dependence with a derived feature importance score allowed to effectively detect model errors.
Detectable errors included: over-fitting, under-fitting, errors caused by imputation, and incorrect cause-effect relationships.
Furthermore, localized inspections helped to understand the how and why of specific instance predictions by finding locally impactful features.

Then, we proposed and discussed the model diagnostic workflow, which is based on aggregating local instance-level explanations in order to gain insights about a model with the intent of improving it.
We showed that the model diagnostic workflow enables a scalable understanding of local decision making of a predictive model and that aggregating instance-level explanations allows for semantic validation of the input data.
As a model can only perform as well as its input data, gaining insights about the limitations of a model in turn help with feature engineering on the input data.

Finally, we showed how histograms can be effectively used to aggregate instance-level explanations.
With the help of comparing meaningful subsets, aggregated instance-level explanations significantly outperformed inspecting individual explanations or aggregation without explanations in detecting biases in the input data while being more scalable than a tabular representation.
Furthermore, we showed that inspecting individual instance-level explanations can be misleading and hurt detecting biases in the input data.

All in all, we demonstrated a workflow for effectively utilizing black-box instance-level explanations in order to improve model correctness via semantic validation.
However, this approach has some limitations to overcome which we will explore next.
Note, that many of those limitations are not necessarily an argument against the presented techniques but merely a starting point for future research.

@@

Black-box analysis methods provide the convenience of using the same algorithm for multiple models.
However, with regards to data some decisions still need to be made.
For example, for the instance-level explanations in \chapref{chap:explainer} we determined that Martens and Provost's~\cite{Martens:2014:EDD:2600518.2600523} algorithm provided consistently better results than Riberio~\etal's\cite{DBLP:journals/corr/RibeiroSG16} algorithm for our use case of highly sparse binary data.
This is due to the fact that different data types require different approaches for explanations.
A binary feature can have an explanation that describes the presence or absence of the feature whereas for a numeric feature the actual value is important.
Our Model Diagnostic Workflow is agnostic to the used instance-level explanation algorithm but this choice still has to be made.

\section{Generalization to Other Forms of Machine Learning}
We only focused on predictive modeling tasks with tabular inputs.
Even though this is a large and popular subset it does not cover the entirety of machine learning or even predictive modeling.
Presented techniques are not immediately transferable to tasks such as online learning or streaming input data, which requires recurrent models, such as LSTM (Long Short Term Memory) models~\cite{Hochreiter:1997:LSM:1246443.1246450}.
Those areas pose further challenges and offer a variety of research opportunities.

>>>>

In this thesis, we discussed the role of visual analytics to explain black-box machine learning.
We saw that global approaches, like feature selection methods, are not informative enough to help understand model decisions.
Different strategies preferred different, equally reasonable, feature sets without having a significant impact on predictive performance.
This showed, that inspecting and comparing alternate settings may let machine learning experts develop insights that overwrite their initial intuitions.

Next, we saw that partial dependence with a derived feature importance score allowed to effectively detect model errors.
Detectable errors included: over-fitting, under-fitting, biases caused by imputation, and leaking labels caused by incorrect cause-effect relationships.
Furthermore, localized inspections helped to understand the how and why of specific instance predictions by finding locally impactful features.

Then, we proposed and discussed the Model Diagnostic workflow, which is based on aggregating local instance-level explanations in order to gain insights about a model with the intent of improving it.
We showed that the Model Diagnostic workflow enables a scalable understanding of local decision making of a predictive model and that aggregating instance-level explanations allows for semantic validation of the input data.
As a model can only perform as well as its input data, gaining insights about the limitations of a model in turn help with feature engineering on said data.

Finally, we showed how histograms can be effectively used to analyze aggregated instance-level explanations.
With the help of comparing meaningful subsets, aggregated instance-level explanations significantly outperformed inspecting individual explanations or aggregation without explanations in detecting biases in the input data while being more scalable than a tabular representation.
Furthermore, we showed that inspecting individual instance-level explanations can be misleading and hurts detecting biases in the input data.

All in all, we demonstrated a workflow for effectively utilizing black-box instance-level explanations in order to improve model correctness via semantic validation.
However, this approach has some limitations to overcome which we will explore next.
Note, that those limitations are not an argument against the presented techniques but merely a starting point for future research.

@@

Black-box analysis methods provide the convenience of using the same algorithm for multiple models.
However, with regards to data some decisions still need to be made.
For example, for the instance-level explanations in \chapref{chap:explainer} we determined that Martens and Provost's~\cite{Martens:2014:EDD:2600518.2600523} algorithm provided consistently better results than Riberio~\etal's\cite{DBLP:journals/corr/RibeiroSG16} algorithm for our use case of highly sparse binary data.
This is due to the fact that different data types require different approaches for explanations.
A binary feature can have an explanation that describes the presence or absence of the feature whereas for a numeric feature the actual value is important.
Our Model Diagnostic workflow is agnostic to the used instance-level explanation algorithm but this choice still has to be made.

\section{Generalization to Other Forms of Machine Learning}
In this thesis, we only focused on predictive modeling tasks with structured inputs.
Even though this is a large and popular subset, it does not cover the entirety of machine learning or even predictive modeling.
Presented techniques are not immediately transferable to tasks such as online learning or streaming input data, which requires recurrent models, such as LSTM (Long Short Term Memory) models~\cite{Hochreiter:1997:LSM:1246443.1246450}.
Those areas pose further challenges and offer a variety of research opportunities.
